---
title: "Dates and Times"
subtitle: "Swirl course notes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```
The package lubridate works with dates and time
"lubridate has a consistent, memorable syntax, that makes working with dates fun instead of frustrating."  
```{r, message=FALSE, warning=FALSE}
require(lubridate)
```
  
```{r}
Sys.getlocale("LC_TIME") # to view the locale of my date&time representation
```
  
###Get and save today's date
```{r}
this_day <- today ()
this_day
```
  
###Components of the date
```{r}
year (this_day)
month (this_day)
day (this_day)
```
  
###Day of the week
```{r}
wday (this_day) # day of the week
wday (this_day, label = T) # print the name of the day
```
  
###day-time combinations
####day & time of the moment
```{r}
this_moment <- now ()
this_moment
```
  
###Components of the day-time combination
```{r}
hour (this_moment)
minute (this_moment)
second (this_moment)
```
<br/>

###Parsing date-times: ``ymd()``, ``dmy()``, ``hms()``, ``ymd_hms()``, etc.  
#### ``ymd()`` took a character string as input and returned an object of class POSIXct  
#### POSIXct is one way that R stores date-time information internally  
```{r} 
my_date <- ymd ("1989-05-17") 
my_date
```

```{r}
class (my_date)
```

```{r}
ymd ("1989 May 17") # different date-time format also works
```

```{r}
mdy ("March 12, 1975") # need to use differnt functions to match the order of components
```

```{r}
dmy (25081985) # works also with numeric strings
```

```{r}
ymd ("192012") # using quotes is consistent with the way most dates are represented (as character strings)
```
  
```{r}
ymd ("1920/1/2") # must supply sufficient info
```

```{r}
dt1 <- "2014-08-23 17:23:02"# Example dataset
```

```{r}
ymd_hms (dt1)
```

```{r}
hms ("03:22:14")
```

```{r}
dt2 <- c("2014-05-14", "2014-09-22", "2014-07-11") # Example dataset
```

```{r}
ymd (dt2)
```
  
### ``update()`` updates 1+ components of a date-time
```{r}
update (this_moment, hours = 8, minutes = 34, seconds = 55) # manually alter the components but NOT the object
```

```{r}
this_moment # remains the same unless I change it
```

```{r}
this_moment <- update (this_moment, hours = 16, minutes = 3) # this alters the object
```

### Switches of time zones
#### Scenario 1: Now, pretend you are in New York City and you are planning to visit a friend in Hong Kong. 
#### You seem to have misplaced your itinerary, but you know that your flight departs New York at 17:34 (5:34pm) the day after tomorrow. 
#### You also know that your flight is scheduled to arrive in Hong Kong exactly 15 hours and 50 minutes after departure.
```{r}
nyc <- now ("America/New_York") # the moment in new york city
nyc
```

for time zone names: <http://en.wikipedia.org/wiki/List_of_tz_database_time_zones>
```{r}
depart <- nyc + days (2) # add 2 days (the day after tomorrow)
depart
```

```{r}
depart <- update (depart, hours = 17, minutes = 34) # departure time
depart
```

```{r}
arrive <- depart + hours (15) + minutes (50) # flight duration
arrive
```

```{r}
arrive <- with_tz(arrive, "Asia/Hong_Kong") # arrival time in hong kong
arrive
```


#### Scenario 2: You and your friend have just met at the airport and you realize that the last time you were together was in Singapore on June 17, 2008.   
#### Naturally, you'd like to know exactly how long it has been.
```{r}
last_time <- mdy ("June 17, 2008", tz = "Singapore")
last_time #"2008-06-17 SGT"
```

``interval (start time, end time...)``ß
```{r}
last_time <- "2008-06-17 SGT" # assign value b/c format not matched
how_long <- interval (last_time, arrive)
as.period (how_long)
```
 