---
title: "data.table how does it work"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library (data.table)
```

## General form
The general form of a ``data.table`` query is ``[i, j, by]``. The 3 parts of the query can be mapped to SQL as _i_ = where, _j_ = select, and _by_ = group by. 
```{r}
mtcarsDT <- data.table (mtcars)
mtcarsDT
mtcarsDT [mpg > 20,
          .(AvgHP = mean(hp),
            "minWT(kg)" = min (wt + 453.6)), # wt in lbs
          by = .(cyl, under5gears = gear < 5)]
```

> Take DT, subset rows using _i_, then calculate _j_ grouped by _by_    
  
In creation of a ``data.table``, use ``L`` after a numeric to specify that it is an integer (NOT _numeric_). Give columns with different lengths, and R will "recycle" the shorter column to match the length of the longer one by re-using the first items
```{r}
## Create a data.table
## Column a will be of integer type
DT <- data.table(a = c(1L, 2L), b = LETTERS[1:4])

## Create another data.table
## Columns B and C will be cycling
DT <- data.table (A = 1:6, B = c("a", "b", "c"), C = rnorm (6), D = TRUE)
```

## Selecting rows by number in _i_
```{r}
DT [3:5,]
DT [3:5] # same thing, unlike data.frame
```

## Compatibility
``data.table`` is a ``data.frame`` too. It is accepted by all base R packages. 
```{r}
class (DT)
```

When a ``data.table`` is called by a function, it looks to see where it is called from. When it is called by a package which does not recognize ``data.table``, it will switch ``data.frame``, and it will still work.    
  
The ``.N`` in ``data.table`` means number of rows
```{r}
DT [.N]
DT [nrow(DT)] # same thing

# Select the second to last row
DT [.N-1]

# Names and dimensions
names (DT)
dim (DT)

# Selective subsetting
DT [c (2, 2, 3)]
```

## Sececting columns in _j_
```{r}
DT [, .(B, C)] # .() is an alias to list() in data.tables and they mean the same
DT [c(1, 3), .(B, C)] # a vector is required for i

# Object or Column names?
B <- 3
DT [, .(B)]
DT [, B]
```

## Computing on columns
Call functions on the columns as if the columns were variables. We can provide new column names while computing. If we don't, column names will be assigned automatically.  
```{r}
DT [, .(Total = sum(A), Mean = mean(C))]

# No column names
DT [, .(sum(A), mean(C))]
```

## Recying in _j_
```{r}
# The value of C will be recycled to fit the length of B
DT [, .(B, C = sum (C))]
```

## Throw anything into J
without even giving a value
```{r}
# A plot and a Null value will be returned
DT [, plot (A, C)]
```

Throw multiple actions in curly brackets
```{r}
DT [, {print(A)
  hist(C)
  NULL}] # set NULL to plot C instead of counting C values
```

## Doing _j_ by group
```{r}
# the groups are returned in order as they appear in the original DT
DT [, .(mySum = sum(A),
        myMean = mean(A)),
    by = .(B)]
```

## Function calls in _by_ 
```{r}
DT [, .(mySum = sum(C)), by = .(Grp = A%%2)]

# if you just have one items each, you can do this:
DT [, sum(C), by = A%%2]
```

## Grouping only on the subset
```{r}
# Note that the even numbers return first because of the selection
DT [2:4, sum(C), by = A%%2]
```

## Practice using iris
```{r}
DT <- data.table (iris)
head (DT)
DT [, mean (Sepal.Length), by = Species]
DT [, mean (Sepal.Length), by = .(substr (Species, 1, 1))]

# .N in j
DT [, .(Count = .N), by = .(Area = round (Sepal.Length* Sepal.Width /10))]
```

## Grouping by multiple columns
```{r}
DT [, .(C = cumsum (Sepal.Length)), .(Petal.Width, Species)]

DT [, .(C = tail (Sepal.Length, 2)), by = .(Species)]
```

## Chaining operations together
```{r}
# in multiple statements
DT <- data.table (A = c("c", "b", "a"), B = 1:6)
DT
ans <- DT [, sum (B), by = A]
ans
ans [order (A)]

# in one sigle single statement
DT [, sum (B), by = A] [order (A)]
```

## Chaining the previous iris example
```{r}
DT <- data.table (iris)
DT [, .(C = cumsum (Sepal.Length)), .(Petal.Width, Species)][, .(C = tail (C, 2)), by = .(Species)]
DT [, .(Sepal.Length = median (Sepal.Length), 
       Sepal.Width = median (Sepal.Width), 
       Petal.Length = median (Petal.Length),
       Petal.Width = median (Petal.Width)), 
    by = Species] [order (Species, decreasing = TRUE)]
```

## Subset of data: .SD 
When you have too many columns and you want to perform particular operation on a subset of columns
```{r}
# .SD represents all columns except those specified in by
DT [, lapply (.SD, median), by = Species]

# .SDcols specifies columns used in .SD
DT [, lapply(.SD, sum), .SDcols =2:4]

# select all but first row of the groups
DT <- data.table (A = c("c", "c", "b", "a"), B = 1:6, C = TRUE, D = 5L)
DT
DT [, .SD[-1], .SDcols = 2:4, by = A]

# sum all cols and number of rows
DT <- data.table (x = c(2,1,2,1,2,2,1),
                  y = c(1,3,5,7,9,11,13),
                  z = c(2,4,6,8,10,12,14))
DT [, c (lapply (.SD, sum), .N), by = x, .SDcols = c ("x", "y", "z")]

# cumsum while grouping
DT [, lapply (.SD, cumsum), .SDcols = 1:2, by = .(by1 = x, by2 = z > 8)]
```

## Add/update columns in _j_ using :=
```{r}
DT
# Add
DT [, c("x", "p") := .(rev(x), 1:7)]
DT
DT [, x := rev(x)] # short expression
DT

# Remove
DT [, c("y", "z") := NULL]
DT [, z := NULL] # short expression
DT

# use objects
myCol <- c ("x", "p")
DT [, (myCol) := .(rev(x), 1:7)] # wrap in (), or it's recognized as column names
DT

# assign multiple columns
DT [, ":=" (y = x*2,
            q = p+1)]
DT
```

## := cimbined with _i_ and _by_
```{r}
DT <- data.table (x = c (2, 2, 1, 1, 1), y = 6:10)
DT [2:4, z := sum (y), by = x]
DT
```

## Add/update multiple columns
```{r}
DT <- data.table (A = c(1, 1, 1, 2, 2), B = 1:5)
DT [,":="(B = B + 1, C = A + B, D = 2)]
DT
```

## Delete by col number
```{r}
DT [, 2 := NULL]
DT
```

## set()
repeated update data by reference
```{r}
DT <- data.table (x = 1:5, y = c (1, 8, 1, 1, 1), z = c (2, 5, 4, 2, 3))
# update the 1:5 rows of the 3rd column into new values
for (i in 1:5) DT [i, z := i + 1]
for (i in 1:5) set (DT, i, 3L, i + 1) # a better way
DT

# set random 3 rows to NA
for (i in 2:3) set(DT, sample(1:5, 3), i, NA)
DT
```

## setnames()
```{r}
setnames (DT, "z", "p")
DT
setnames (DT, names (DT), toupper (names(DT)))
DT
```

## setcolorder()
```{r}
setcolorder (DT, c("P", "X", "Y"))
DT
```

## The ``set`` family
```{r}
setnames (DT, names (DT), paste0 (names (DT), "_2"))
DT
setcolorder (DT, rev (names (DT)))
DT
```

## Using column names in i
```{r}
DT <- data.table (A = c("c", "b", "a"), B = 1:6)
DT
DT [A == "a"]
DT [A %in% c ("a", "c")]

# Indexing
w <- DT [, A == "a"] # return a vector if not wrapped in .()
w
DT [w]
```

## Indexing using iris
```{r}
DT <- data.table(iris)

# Species is "virginica"
DT [Species == "virginica"]

# Species is either "virginica" or "versicolor"
DT [Species %in% c("virginica", "versicolor")]

# Remove the "Sepal." prefix
setnames (DT, gsub("^Sepal\\.", "", names (DT)))
DT

# Remove the two columns starting with "Petal"
DT [, grep ("^Petal", names (DT)) := NULL]
DT
```

## Automatic indexing  
"if i is a single variable name, it is evaluated in the calling scope, otherwise inside DT's scope" Only single columns on the left side of operators benefit from automatic indexing.  
```{r}
# Cleaned up iris data.table
DT <- data.table(iris)

# Area is greater than 20 square centimeters
DT [ Sepal.Width * Sepal.Length > 20 ]

# Add new boolean column
DT [, is_large := Sepal.Width * Sepal.Length > 25]
DT

# Now large observations with is_large
DT [is_large == TRUE]
DT [(is_large)] # Also OK
```

## Creating and using a key
When a key is being set, the data.table sorts the data by the column we specify and change the table by reference. Having set a key will allow you to use it, for example, as a super-charged row name when doing selections.   
```{r}
DT <- data.table (iris [c(101:103, 51:53, 1:3),])
DT
setkey (DT, Species)
DT

# indexing
DT ["setosa"]

# Arguments like mult and nomatch then help you to select only particular parts of groups. 
DT ["setosa", mult = "first"] # the first row of selection
DT ["setosa", mult = "last"] # the last row of selection

# Indexing with no match item
DT [c ("setosa", "gropa")] # default is nomatch = NA
DT [c ("setosa", "gropa"), nomatch = 0]
```

## Two-column key
The second key is sorted within each group of the first key
```{r}
setkey (DT, Species, Sepal.Length)
DT

# indexing both keys
DT [.("setosa", 4.7)]

# indexing with no match item
DT [.("setosa", 5.2)]

# indexing with only the first key
DT [.("setosa")]
```

## See the difference of indexing with and without a key
```{r}
DT <- data.table (iris [c (101:103, 51:53, 1:3),])
DT
DT [Species == "setosa"] # without key
setkey (DT, Species)
DT ["setosa"] # with key
```

## Selecting groups or parts of groups
```{r}
# The 'keyed' data.table DT
DT <- data.table(A = letters[c(2, 1, 2, 3, 1, 2, 3)], 
                 B = c(5, 4, 1, 9, 8, 8, 6), 
                 C = 6:12)
setkey(DT, A, B)

# Select the "b" group
DT ["b"]

# "b" and "c" groups
DT [c ("b", "c")]

# The first row of the "b" and "c" groups
DT [c ("b", "c"), mult = "first"]

# First and last row of the "b" and "c" groups
DT [c ("b", "c"), .SD[c(1, .N)], by = .EACHI]

# Copy and extend code for instruction 4: add printout
DT [c("b", "c"), { print(.SD); .SD[c(1, .N)] }, by = .EACHI]
```

## Rolling joins
```{r}
DT <- data.table (iris [c (48:53, 101:103), ])
setkey (DT, Species, Sepal.Length)
DT

DT [.("setosa", 4.9)] # there is no match of 4.9 in the second key = a gap
DT [.("setosa", 4.9), roll = TRUE] # allow to roll the prevailing observation before the gap forward = show other columns across the gap

DT [.("setosa", 4.9), roll = "nearest"] # the nearest value that match the second key

# Always roll forwards or backwards
DT [.("setosa", 4.9), roll = +Inf]
DT [.("setosa", 4.9), roll = -Inf]

# Limit the staleness of data
DT [.("setosa", 4.9), roll = +2]
DT [.("setosa", 4.9), roll = -2]

# Concrol ends
DT [.("setosa", 6:7), roll = TRUE]
DT [.("setosa", 6:7), roll = TRUE, rollends = FALSE]
```


## Practice
```{r}
# Keyed data.table DT
DT <- data.table(A = letters[c(2, 1, 2, 3, 1, 2, 3)], 
                 B = c(5, 4, 1, 9, 8, 8, 6), 
                 C = 6:12, 
                 key = "A,B")

# Get the key of DT
key (DT)

# Row where A == "b" and B == 6
DT[.("b",6)]

# Return the prevailing row
DT[.("b",6), roll = TRUE]

# Return the nearest row
DT[.("b",6), roll = "nearest"]
```


The rollends argument is actually a vector of two logical values, but remember that you can always look this up via ?data.table. You were introduced to this argument via the control ends section. If you want to roll for a certain distance, you should continue to use the roll argument.
```{r}
# Keyed data.table DT
DT <- data.table(A = letters[c(2, 1, 2, 3, 1, 2, 3)], 
                 B = c(5, 4, 1, 9, 8, 8, 6), 
                 C = 6:12, 
                 key = "A,B")

# Print the sequence (-2):10 for the "b" group
DT[.("b", (-2):10)]

# Add code: carry the prevailing values forwards
DT[.("b", (-2):10), roll = TRUE]

# Add code: carry the first observation backwards
DT[.("b", (-2):10), roll = TRUE, rollends = TRUE]
DT[.("b", (-2):10), roll = TRUE, rollends = c(TRUE, TRUE)] # also OK
```
