---
title: "Tidying data with tidyr"
subtitle: "swirl course notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r message=FALSE, warning=FALSE}
library (tidyr); library (readr)
```
**Tidy data** is formatted in a standard way that facilitates exploration and
analysis and works seamlessly with other tidy data tools. Specifically,
tidy data satisfies three conditions:

1. Each variable forms a column  
2. Each observation forms a row  
3. Each type of observational unit forms a table  

Any dataset that doesn't satisfy these conditions is considered '**messy**' data.  
Therefore, all of the following are characteristics of messy data,  
EXCEPT...6  
1. Column headers are values, not variable names  
2. Variables are stored in both rows and columns  
3. A single observational unit is stored in multiple tables  
4. Multiple types of observational units are stored in the same table  
5. Multiple variables are stored in one column  
6. **Every column contains a different variable**   

### Problem 1: column headers are values, not variable names  
#### sample dataset #1
```{r}
students <- read.table (file = "../data/students.txt", header = T)
students
```
The first column represents each of five possible grades that students could receive for a particular class.  
The second and third columns give the number of male and female students, respectively, that received each grade.  
This dataset actually has 3 variables: grade, sex, and count.  
The first variable, **grade**, is already a column, so that should remain as it is.  
The second variable, **sex**, is captured by the second and third column headings.  
The third variable, **count**, is the number of students for each combination of grade and sex.  

#### To tidy the students data, we need to have one column for each of these 3 variables.  
```{r}
# EXCEPT the grade column
gather(students, sex, count, -grade)
```
Each row of the data now represents exactly one observation, characterized by a unique combination of the grade and sex variables.  
Each of our variables (grade, sex, and count) occupies exactly one column. That's tidy data!  

***  
### Problem 2: multiple variables are stored in one column.  
#### sample dataset #2
```{r}
students2 <- read.table (file = "../data/students2.txt", header = T)
students2
```
This dataset is similar to the first, except now there are two separate classes, 1 and 2, and we have total counts for each sex within each class.  
students2 suffers from the same messy data problem of having column headers that are values (male_1, female_1, etc.) and not variable names (sex, class, and count).  
However, it also has multiple variables stored in each column (sex and class), which is another common symptom of messy data.  

#### Tidying this dataset will be a two step process.
```{r}
# STEP 1: stack the columns
res <- gather (students2, sex_class, count, -grade)
res
```
That got us half way to tidy data, but we still have two different variables, sex and class, stored together in the sex_class column.  
tidyr offers a convenient ``separate()`` function for the purpose of separating one column into multiple columns.  
```{r}
# STEP 2: split the sex_class column into sex and class
separate (data = res, col = sex_class, into = c("sex", "class"))
```
Conveniently, ``separate()`` was able to figure out on its own how to separate the sex_class column.  
**Unless you request otherwise with the 'sep' argument, it splits on non-alphanumeric values.**  
In other words, it assumes that the values are separated by something other than a letter or number (in this case, an underscore.)  

#### combine the 2 steps
```{r}
students2 %>%
  gather (sex_class, count, -grade) %>%
  separate (sex_class, c("sex", "class")) %>%
  print
```
***
### Problem 3: variables are stored in both rows and columns.  
#### sample dataset #3  
```{r}
students3 <- read.table (file = "../data/students3.txt", header = T)
students3
```
In students3, we have midterm and final exam grades for five students, each of whom were enrolled in exactly two of five possible classes.  
The first variable, **name**, is already a column and should remain as it is.  
The headers of the last five columns, **class1** through **class5**, are all different values of what should be a class variable.  
The values in the **test** column, midterm and final, should each be its own
variable containing the respective grades for each student.  

#### This will require multiple steps
```{r message=FALSE, warning=FALSE}
# STEP 1: gather the columns class1 through class5 into a new variable called class
students3 %>%
  gather (class, grade, class1:class5, na.rm = TRUE) %>%
  # Use the argument na.rm = TRUE to omit NA values from the final result
  print
```
tidyr makes it easy to reference multiple adjacent columns with ``class1:class5``, just like with sequences of numbers.  
Since each student is only enrolled in two of the five possible classes, there are lots of missing values (i.e. NAs).  
```{r message=FALSE, warning=FALSE}
# STEP 2: turn the values of the test column, midterm and final, into column headers (i.e. variables)
students3 %>%
  gather(class, grade, class1:class5, na.rm = TRUE) %>%
  spread(test, grade) %>%
  print

# STEP 3: we want the values in the class column to simply be 1, 2, ..., 5  and not class1, class2, ..., class5
library(readr)
parse_number("class5")
 
# STEP 4: "overwrite" a column with mutate() by assigning a new value to the existing column instead of creating a new column
library(dplyr)
students3 %>%
  gather(class, grade, class1:class5, na.rm = TRUE) %>%
  spread(test, grade) %>%
  mutate(class = parse_number(class)) %>%
  print
```
***  
### Problem 4: multiple observational units are stored in the same table.  
#### sample dataset #4  
```{r}
students4 <- read.table (file = "../data/students4.txt", header = T)
students4
```
students4 is almost the same as our tidy version of students3.  
The only difference is that students4 provides a unique **id** for each student, as well as his or her **sex** (M = male; F = female).  
At first glance, there doesn’t seem to be much of a problem with students4.  
All columns are variables and all rows are observations.  
However, notice that each **id**, **name**, and **sex** is repeated twice, which seems quite redundant.  
This is a hint that our data contains multiple observational units in a single table.  
#### Our solution will be to break students4 into two separate tables – one containing basic student information (**id**, **name**, and **sex**) and the other containing grades (**id**, **class**, **midterm**, **final**).  
```{r}
# STEP 1: select the id, name, and sex column from students4 and storing the result in student_info
student_info <- students4 %>%
  select(id, name, sex) %>%
  print
# STEP 2: remove duplicate rows from student_info.  
student_info <- students4 %>%
  select(id, name, sex) %>%
  unique %>%
  print
# STEP 3: select the id, class, midterm, and final columns (in that order) and store the result in gradebook.  
gradebook <- students4 %>%
  select(id, class, midterm, final) %>%
  print
```
It's important to note that **we left the id column in both tables**.  
**In the world of relational databases, 'id' is called our'primary key' since it allows us to connect each student listed in student_info with their grades listed in gradebook.**  
Without a unique identifier, we might not know how the tables are related.  
(In this case, we could have also used the name variable, since each student happens to have a unique name.)  

***  
### Problem 5: a single observational unit is stored in multiple tables  
### (It's the opposite of the fourth problem)  
#### sample datasets #5
```{r}
passed <-read.table (file = "../data/passed.txt", header = T)
passed
failed <-read.table (file = "../data/failed.txt", header = T)
failed
```
Teachers decided to only take into consideration final exam grades in determining whether students passed or failed each class. 
As you may have inferred from the data, students passed a class if they received a final exam grade of A or B and failed otherwise.  
The name of each dataset actually represents the value of a new variable that we will call '**status**'.  
Before joining the two tables together, we'll add a new column to each containing this
information so that it's not lost when we put everything together.  
```{r}
passed <- passed %>%
  mutate (status = "passed")
passed
failed <- failed %>%
  mutate (status = "failed")
failed
```
Now, pass as arguments the passed and failed tables (in order) to the dplyr function ```bind_rows()```, which will join them together into a single unit.  
```{r message=FALSE, warning=FALSE}
bind_rows(passed, failed)
```
Of course, we could arrange the rows however we wish at this point, but **the important thing is that each column is a variable, and the table contains a single observational unit**.  
Thus, the data are tidy.  

***  
### Bring everything together
#### the 2013 SAT data 
```{r}
sat <- read.table(file = "../data/sat.txt", header = T)
sat
```
Accomplish the following three goals:  

1. ``select()`` all columns that do NOT contain the word "**total**", since if we have the male and female data, we can always recreate the total count in a separate column, if we want it.  
Hint: Use the ``contains()`` function, which you'll find detailed in 'Special functions' section of ``?select``.  
2. ``gather()`` all columns EXCEPT **score_range**, using ``key = part_sex`` and ``value = count``.  
3. ``separate()`` **part_sex** into two separate variables (columns), called "**part**" and "**sex**", respectively.  You may need to check the 'Examples' section of ``?separate`` to remember how the '``into``' argument should be phrased.  
```{r}
sat %>%
  select(-contains("total")) %>%
  gather(part_sex, count, -score_range) %>%
  separate(part_sex, c("part", "sex")) %>%
  print
```
Append two more function calls to accomplish the following:  

1. Use ``group_by()`` (from dplyr) to group the data by **part** and **sex**, in that order.  
2. Use ``mutate`` to add two new columns, whose values will be automatically computed group-by-group:  
  
  * ``total = sum(count)``  
  * ``prop = count / total``  
```{r}
sat %>%
  select(-contains("total")) %>%
  gather(part_sex, count, -score_range) %>%
  separate(part_sex, c("part", "sex")) %>%
  group_by(part, sex) %>%
  mutate(total = sum(count),
         prop = count / total
  ) %>% print
```
In this lesson, you learned how to tidy data with tidyr and dplyr.  
These tools will help you spend less time and energy getting your data ready to analyze and more time actually analyzing it.  